Designing Types for R, Empirically

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd
Update static type analysis
Add graph for unions

---
title: Evaluation of Types
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
params:
  type_analysis_datafile: ../data/type-analysis/merged/merged.csv
---

@@

@@

```{r incompatible-functions, echo=FALSE, eval=TRUE}
type_compatibility <-
    type_analysis_data %>%
    group_by(package_name, function_name, outer_alternative) %>%
    summarize(parameter_count = length(unique(parameter_position))) %>%
    ungroup() %>%
    group_by(package_name, function_name) %>%
    mutate(compatible = length(unique(parameter_count)) == 1) %>%
    ungroup()

@@

union_frequency <-
union_data %>%
distinct(package_name, function_name, parameter_position, alternative_count) %>%
group_by(alternative_count) %>%
summarize(parameter_count = n()) %>%
ungroup() %>%
arrange(desc(parameter_count)) %>%
mutate(percentage = round(100 * parameter_count / sum(parameter_count), 2),
       cumulative_percentage = round(100 * cumsum(parameter_count) / sum(parameter_count), 2))

@@

@@

datatable(union_types)
```

>>>>

---
title: Evaluation of Types
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
params:
  type_analysis_datafile: ../data/type-analysis/merged/merged.csv
  graph_dirpath: ../../paper/plots/
---

@@

source("latex-macros.R")

save_graph <- function(plot, filename, device = cairo_pdf, width=12, height=4, ...) {
    ggsave(filename = file.path(params$graph_dirpath, filename), plot = plot, device = device, width=width, height=height, ...)
    plot
}

@@

```{r incompatible-functions, echo=FALSE, eval=TRUE}

###type_compatibility <-
###    type_analysis_data %>%
###    group_by(package_name, function_name, outer_alternative) %>%
###    summarize(parameter_count = length(unique(parameter_position))) %>%
###    ungroup() %>%
###    group_by(package_name, function_name) %>%
###    mutate(compatible = length(unique(parameter_count)) == 1) %>%
###    ungroup()

type_compatibility <-
    type_analysis_data %>%
    group_by(package_name, function_name) %>%
    summarize(compatible = length(unique(outer_alternative)) == 1) %>%
    ungroup()

@@

union_frequency <-
union_data %>%
distinct(package_name, function_name, parameter_position, alternative_count) %>%
group_by(alternative_count) %>%
summarize(parameter_count = n()) %>%
ungroup() %>%
arrange(desc(parameter_count)) %>%
mutate(percentage = round(parameter_count / sum(parameter_count), 4),
       cumulative_percentage = round(cumsum(parameter_count) / sum(parameter_count), 4))

@@

total_parameter_count <- sum(union_frequency$parameter_count)

print(total_parameter_count)

union_frequency_folded <-
    union_frequency %>%
    select(alternative_count, parameter_count, percentage) %>%
    filter(alternative_count > 3) %>%
    summarize(alternative_count = "> 3", parameter_count = sum(parameter_count)) %>%
    mutate(percentage = round(as.double(parameter_count) / total_parameter_count, 3)) %>%
    print()

union_frequency_folded <-
    union_frequency %>%
    select(alternative_count, parameter_count, percentage) %>%
    filter(alternative_count <= 3) %>%
    mutate(alternative_count = as.character(alternative_count)) %>%
    rbind(union_frequency_folded)

datatable(union_frequency_folded)

(union_frequency_folded %>%
ggplot(aes(alternative_count, percentage)) +
geom_col() +
scale_y_continuous(sec.axis = sec_axis(~ . * total_parameter_count, labels = count_labels), labels = relative_labels) +
labs(x="Types", y="Parameters") +
scale_x_discrete(limits=c("> 3", 3, 2, 1)) +
coord_flip() +
theme(legend.title = element_blank(), legend.position = "top")) %>%
save_graph("union_frequency.pdf")

@@

datatable(union_types)

single_union_types <-
union_data %>%
filter(alternative_count == 1) %>%
group_by(kind) %>%
summarize(count = n())%>%
ungroup() %>%
arrange(desc(count)) %>%
mutate(percentage = round(100 * count / sum(count), 1),
       cumsum_percentage = round(100 * cumsum(count) / sum(count), 1))

datatable(single_union_types)
```

### 2020-05-15 paper/main.tex
Add graph for unions

\subsection{Is our simple type language expressive enough to type R code?} Functions in
R can support arguments of multiple types by reflecting on the type to
perform appropriate conversions and computations. On the one hand, this form
of ad-hoc polymorphism makes R functions expressive; on the other hand, it
requires support for polymorphism in the typing scheme, making the type
system complex.  Our system lacks explicit support for
polymorphism. Instead, it supports union and any types. A union type is
inferred when values of multiple types are observed at a function parameter
and return position. any type is inferred when an argument is missing or
unused. Furthermore, list element types can be inferred as any if they have
more than 5 different types. Values with many different types can result in
a large number of alternatives in the type inferred for that position,
making it hard to read and unwieldy for end-users. To evaluate this
''expressivity gap`' between our type inference scheme and R functions, we
look at the number of different types in the parameter and return types of
the XXX functions in our corpus.

>>>>

\subsection{Is our simple type language expressive enough to type R code?} Functions in
R can support arguments of multiple types by reflecting on the type to
perform appropriate conversions and computations. On the one hand, this form
of ad-hoc polymorphism makes R functions expressive; on the other hand, it
requires support for polymorphism in the typing scheme, making the type
system complex.  Our system lacks explicit support for
polymorphism. Instead, it supports union, any and ... types. A union type is
inferred when values of multiple types are observed at a function parameter
and return position, any type is inferred when an argument is missing or
unused, and ... is inferred for vararg parameters. Furthermore, list element
types can be inferred as any if they have more than 5 different types. Values with many different types can result in
a large number of alternatives in the type inferred for that position,
making it hard to read and unwieldy for end-users. To evaluate this
''expressivity gap`' between our type inference scheme and R functions, we
look at the number of different types in the parameter and return types of
the XXX functions in our corpus.

\begin{figure}[!h]  \centering
  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
  \caption{Types inferred for each position}\label{fig:unionfreq}
\end{figure}
